\documentclass[11pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage{amsmath}
\usepackage{hyperref}

\title{Beyond Matmul: Provenance-Aware Computation for Fixed-Weight ML Inference}
\author{Beyond Matmul Contributors}
\date{Final draft -- July 2026}

\begin{document}

\maketitle

\begin{abstract}
Modern machine-learning systems routinely lower semantically rich computations
into dense matrix multiplication because dense GEMM is universal, convenient,
and heavily optimized. That default is useful, but it can erase why a matrix
exists: a convolution becomes an anonymous Toeplitz matrix, an adapter becomes a
materialized product, an embedding projection becomes a dense multiply, and a
quantized operator becomes dequantized storage. This paper explores a
provenance-aware framing for fixed-weight inference in which dense GEMM remains
a valid fallback rather than the default semantic intermediate representation.
We propose a linear and affine operator IR that preserves or recovers structure,
a planner that selects exact or bounded-error lowerings under backend and reuse
contracts, and an evaluation program that ties frontend coverage, tests,
benchmarks, and case studies to the research claims.
\end{abstract}

\section{North Star}

The north star of Beyond Matmul is a completed research artifact with two
mutually reinforcing outputs:

\begin{enumerate}
  \item A final whitepaper that explains, evaluates, and bounds the claim that
  provenance-aware computation can improve fixed-weight ML inference by
  retaining structure normally lost during dense lowering.
  \item An accompanying codebase that provides executable evidence: operator
  semantics, Torch frontend capture, recovery probes, approximation builders,
  planner decisions, tests, demos, and benchmark data.
\end{enumerate}

The project is complete when the paper's claims are supported by the codebase
and its generated evidence, the GitHub wiki concisely defines the project for
humans and agents, and GitHub issues no longer contain unresolved high-priority
gaps against the stated research questions.

After the first artifact, the PEFT plus Transformers capstone moved the project
from controlled evidence to an external PyTorch inference stack. The
TorchBench-style artifact from issue \#80 and PR \#90 records upstream PEFT,
upstream merged dense, and Beyond Matmul PEFT-fork rows for a pinned
fixed-LoRA causal-LM workload. It is useful evidence for provenance visibility:
successful fork rows emit structured LoRA events while dense fallback remains
available. It is negative evidence for performance: the seq128 rows fail across
all baselines, \texttt{summary.benchmark\_ready} is false, and
\texttt{summary.performance\_claim} is \texttt{none}.

\section{Problem}

Dense matrix multiplication is an excellent implementation target and a robust
fallback abstraction. It is not always an excellent semantic representation.
When a framework lowers a structured computation into an anonymous matrix, it
often discards facts that matter for inference:

\begin{itemize}
  \item convolution preserves local receptive fields and weight sharing;
  \item low-rank adapters preserve factorization and rank bounds;
  \item embeddings preserve lookup semantics and repeated token structure;
  \item masks preserve causal, banded, block, or page sparsity;
  \item quantized weights preserve packing, scale, zero point, and codebooks;
  \item diagonal gates preserve elementwise scaling rather than general mixing.
\end{itemize}

Once these facts are erased, a downstream compiler or runtime must either use
dense GEMM or rediscover structure heuristically from the dense matrix. The
first path can waste work and memory; the second path is lossy and expensive to
trust.

\section{Thesis}

Many matrix multiplications in ML systems are semantically structured
computations that were densified for convenience. A provenance-aware
linear-operator IR can preserve or recover that structure and allow a planner to
choose exact or bounded-error lowerings that are cheaper, clearer, or more
controllable than dense GEMM for fixed-weight inference.

The claim is deliberately not ``GEMM is bad.'' Dense GEMM should remain one
lowering in a richer operator space. The research question is when and how much
additional provenance lets the system choose something better.

\section{Scope}

The initial artifact focuses on fixed-weight inference. This scope makes reuse
contracts explicit: preprocessing can be amortized across calls, weights do not
mutate during training, and exact dense equivalence can be checked for small
operators.

Out of scope for the first complete artifact:

\begin{itemize}
  \item training-time weight mutation and gradient semantics;
  \item production GPU kernels;
  \item claims of hardware-calibrated speedup without benchmark evidence;
  \item frontend capture of quantized Torch modules beyond the tested
  per-tensor affine fixed-weight linear contract;
  \item universal Torch operation support in a single pass.
\end{itemize}

\section{Computation Reason Provenance}

This project uses provenance to mean the reason an operator exists, not merely
where the tensor was allocated. A provenance-aware operator records:

\begin{itemize}
  \item source framework node or expression;
  \item input and weight identities;
  \item transformation history, such as convolution to Toeplitz or adapter
  factors to a merged dense weight;
  \item structural metadata, such as rank, sparsity, kernel shape, codebook, or
  quantization layout;
  \item exactness or approximation contract;
  \item reuse and cache assumptions;
  \item backend and layout constraints.
\end{itemize}

This framing lets later passes ask not only ``what matrix is this?'' but ``what
computation did this matrix come from, and what lowerings are still valid?''

\section{System Design}

The artifact is organized around six components.

\subsection{Frontend Capture}

The first method is to capture fixed-weight computation before the framework
has lowered it to an anonymous dense product. The current Torch FX frontend
walks graph nodes, resolves fixed parameters, buffers, constants, and selected
module attributes, then emits executable IR operators whose provenance records
the source node, expression, transform history, and captured structure. This is
a provenance-preserved path: the structured operator is produced from the
original computation rather than inferred from a dense matrix after the fact.

The supported surface is intentionally narrow and test-backed. The frontend
currently captures nested \texttt{nn.Linear} or \texttt{F.linear} factors,
named adapter pairs, embedding followed by projection, single-channel and
multi-channel \texttt{Conv1d} including grouped/depthwise and scalar
stride/padding/dilation variants, functional \texttt{conv1d}, fixed-weight
\texttt{matmul}/\texttt{mm} forms, fixed-bias \texttt{addmm}, and exported
graph fixed-weight \texttt{addmm}/nested-linear recovery through graph
signature state, and fixed per-tensor affine quantized \texttt{nn.Linear}
modules. It does not claim \texttt{Conv2d}, quantized convolutional modules,
per-channel/per-axis or dynamic quantized modules, non-scalar Conv1d
parameters, invalid Conv1d output lengths, non-default \texttt{addmm} scaling,
shape-inferred matmul orientation, or dynamic-weight matmul/addmm. The coverage
matrix currently has supported and unsupported rows, with no intermediate
\texttt{Next} rows; broader exported graph variants and unsupported quantized
operators remain future work until represented and tested. The evidence anchor
is the \texttt{Torch frontend coverage} row in
\texttt{docs/evidence\_matrix.md}, with the detailed boundary maintained in
\texttt{docs/torch\_frontend\_coverage.md}.

\subsection{Operator IR and Contracts}

The IR represents fixed-weight linear maps with shape
$(\mathrm{out\_features}, \mathrm{in\_features})$ and affine maps as a linear
operator plus bias. Applying a linear operator to a row-major batch computes
$xW^T$; applying an affine operator computes $xW^T + b$. Metadata stores
provenance, structural fields, approximation contract, quantization placeholder
data, reuse budget, layout, hardware target, and allowed lowerings. The
contract is part of the operator rather than an external note: exact operators
require backend and layout support, while approximate operators additionally
carry metric, epsilon, observed error, and sample-count evidence.

Implemented operator families include dense, diagonal, sparse COO, fixed-mask,
low-rank, affine, codebook, bitpacked binary, packed affine quantized,
single-channel Conv1d, and multi-channel Conv1d operators. Fixed-mask operators
cover explicit binary masks applied as sparse linear maps over values or
features, independent of input-dependent attention scores. Conv1d operators
keep stride, padding, dilation, output length, and grouped/depthwise channel
partition metadata so the planner can distinguish direct convolution lowerings
from dense materialization. Every
implemented exact family exposes executable semantics and a dense fallback,
which keeps dense GEMM available for equivalence checks and portability. The
evidence anchor is the \texttt{IR operator families} row in
\texttt{docs/evidence\_matrix.md}; \texttt{docs/ir\_spec.md} is the compact
schema reference.

The quantized fixed-weight contract is deliberately narrow. Codebook operators
preserve integer codes plus an explicit value table; bitpacked binary operators
preserve one sign bit per weight plus a tensor-wide scale. These operators may
be exact source representations or approximate candidates depending on their
approximation contract. Dense dequantization remains a valid fallback, but it
is not by itself evidence that packed or codebook provenance survived. The IR
metadata can describe per-tensor scale and zero point, but per-tensor affine
integer payloads now preserve integer storage, bit width, integer range, scale,
and zero point through \texttt{PackedAffineQuantizedOperator}. The frontend only
claims the tested fixed per-tensor affine quantized \texttt{nn.Linear} module
case; per-channel or per-axis quantization, mixed precision,
activation-dependent quantization, and quantized convolution remain future work
until payload, layout, and frontend tests exist.

\subsection{Recovery Analyzer}

When provenance is lost, cheap analyzers probe dense matrices for recoverable
structure. This is the dense-recovery path, and it is intentionally weaker
than frontend preservation. A recovered candidate may explain a matrix, but it
cannot prove the original computation reason unless separate provenance is
available. For that reason, analyzer output carries bounded heuristic
confidence, explicit probe details, and, when sample inputs are supplied,
output-relative validation evidence.

The current probes cover diagonal residuals, sparsity, block sparsity,
permutation-like structure, small codebooks, low-rank approximation, repeated
blocks, and repeated fixed-weight sightings. Confidence values are
probe-specific ranking scores, not calibrated probabilities. If a recovered
candidate is executable, validation on representative inputs records
\texttt{output\_relative\_l2}, sample count, sample exactness, and a
validation-derived confidence cap. Dense GEMM remains the exact fallback
whenever recovery is ambiguous or unsupported. The evidence anchor is the
\texttt{Recovery after lost provenance} row in
\texttt{docs/evidence\_matrix.md}.

\subsection{Approximation Builders}

Approximation builders produce candidate low-rank, sparse, codebook, or binary
operators from a dense reference. The method is deliberately candidate-based:
builders create possible structured operators, attach approximate contracts,
and leave acceptance to validation and planning rather than treating matrix
compression as sufficient evidence. Supported builders currently include
low-rank approximation, sparse top-k selection, codebook quantization, and
bitpacked binary approximation.

Approximate candidates are scored with product-aware metrics when
representative inputs are available. Matrix-relative Frobenius error remains
useful for reconstruction, but output-relative error can reject a candidate
that looks good as a matrix approximation and behaves poorly on the inference
inputs of interest. The current deterministic ablation is intentionally small:
one synthetic dense case shows low-rank and sparse top-k candidates passing a
matrix-relative threshold but failing the same output-relative threshold, while
codebook and bitpacked candidates fail both. This supports the need for
output-aware acceptance; it is not broad model-quality evidence. The evidence
anchor is the \texttt{Approximation and error contracts} row in
\texttt{docs/evidence\_matrix.md}.

\subsection{Planner Selection}

The planner selects a valid lowering under exactness, error, reuse, backend,
and layout contracts. It enumerates the original operator, valid structured or
approximate candidates, and a dense fallback, rejects options whose contracts
do not satisfy the request, estimates cost proxies for the remaining options,
and selects the lowest-cost valid plan. Lowering names include direct Conv1d
variants, low-rank product, sparse kernel, codebook kernel, bitpacked kernel,
and dense GEMM or dense GEMM with bias.

The request specifies exact-only or bounded-error behavior, epsilon and error
metric, representative sample inputs for product-aware scoring, expected call
count for reuse amortization, and backend/layout constraints. Planner costs are
currently estimates: operation count, memory movement, cache footprint,
preprocessing cost, and amortized preprocessing over repeated calls. The
planner-contract ablation is therefore contract evidence rather than hardware
performance evidence. It compares exact-only and bounded-error requests on the
same fixed-weight case, rejects and then accepts a preprocessing-heavy
low-rank lowering across its reuse threshold, and rejects an unsupported GPU
codebook lowering while preserving dense GEMM as a valid fallback. The evidence
anchor is the \texttt{Planner exactness and fallback} row in
\texttt{docs/evidence\_matrix.md}.

\subsection{Evidence Surface}

The codebase must produce evidence that the paper can cite: tests for semantics
and negative capture cases, workload case-study artifacts for adapter, Conv1d,
fixed-mask, and quantized-linear narratives, benchmark artifacts for cost
comparisons, an approximation-error ablation artifact, and docs/wiki pages that
describe the current boundary of truth. The support material includes a concise
evidence matrix in \texttt{docs/evidence\_matrix.md} that maps major claims to
tests, demos, benchmark artifacts, or future-work boundaries.

\section{Research Questions}

\begin{enumerate}
  \item Which common fixed-weight Torch inference patterns can be captured
  exactly before they become anonymous dense matrices?
  \item When provenance is missing, which structures can be recovered cheaply
  enough to be useful, and how should uncertainty be represented?
  \item Under what reuse and backend contracts does a structured lowering beat
  dense GEMM according to operation, memory, preprocessing, or measured latency
  proxies?
  \item How should approximate lowerings be evaluated so output-level error is
  aligned with real inference behavior?
  \item What minimum wiki, issue, and automation loop keeps agentic development
  aligned with the research goal instead of accumulating unrelated code?
\end{enumerate}

\section{Evaluation Plan}

The evaluation should combine controlled synthetic cases and small real
workload case studies.

\subsection{Controlled Cases}

\begin{itemize}
  \item exact diagonal operators;
  \item exact sparse operators;
  \item exact low-rank operators;
  \item exact convolutional operators;
  \item small-codebook and bitpacked approximations;
  \item random dense fallback baselines.
\end{itemize}

\subsection{Workload Case Studies}

\begin{itemize}
  \item Conv1d module and functional Conv1d capture before dense
  materialization;
  \item LoRA or adapter weights merged into a dense projection;
  \item PEFT plus Transformers inference with fixed LoRA/adapters, comparing
  upstream PEFT against a Beyond Matmul PEFT fork under a TorchBench-style
  reproducible harness. The first measured artifact is
  \texttt{docs/results/peft\_transformers\_lora\_inference.json}, produced for
  issue \#80 and merged in PR \#90 under the contract recorded in
  \texttt{docs/peft\_capstone\_benchmark\_contract.md};
  \item Conv/linear lowering from a broader CNN-style block;
  \item fixed binary masks applied as sparse linear maps, with full masked
  attention left as future work because score computation, softmax, broadcast
  mask contracts, KV-cache layout, and dynamic sequence behavior sit outside
  the current fixed-weight linear boundary;
  \item attention projection with fixed structured weights;
  \item embedding plus projection in a language-model-like block;
  \item quantized linear modules for the tested per-tensor affine case, with
  quantized convolution left for future packed-payload rules.
\end{itemize}

\subsection{Metrics}

\begin{itemize}
  \item dense equivalence for exact lowerings;
  \item output-relative error for approximate lowerings;
  \item matrix-reconstruction versus output-error decisions for a bounded
  approximation ablation;
  \item operation-count proxy;
  \item memory-footprint and memory-movement proxy;
  \item preprocessing cost and calls required to amortize it;
  \item wall-clock latency where measurement is meaningful;
  \item coverage matrix rows connected to executable tests.
\end{itemize}

\subsection{Current Benchmark Results}

The current results are generated artifacts rather than hand-curated tables.
The main fixed-weight benchmark is regenerated with
\texttt{mise exec -- uv run python benchmarks/fixed\_weight.py --json-output
docs/results/fixed\_weight.json}; local CI also regenerates the workload
case-study, approximation-error, and planner-contract JSON artifacts documented
in \texttt{docs/benchmark\_artifacts.md}.

In the fixed-weight benchmark, the planner request allows approximate
lowerings with a maximum relative error of $0.05$ over a batch of $32$ and
$32$ requested calls. The current artifact selects exact specialized lowerings
for diagonal, sparse, low-rank, small-codebook, single-channel Conv1d, and
multi-channel Conv1d cases, and selects dense GEMM for the random dense
fallback case. All rows are valid and exact in this controlled benchmark. The
recorded fields include selected lowering, relative error, estimated apply
cost, estimated memory bytes, preprocessing cost, amortized preprocessing, and
pure-Python seconds per apply for dense and selected lowerings. For example,
the diagonal and sparse cases select \texttt{diagonal\_kernel} and
\texttt{sparse\_kernel} with zero output-relative error and no preprocessing,
while the codebook case records preprocessing work that is amortized over the
requested calls. The Conv1d rows select direct Conv1d lowerings with zero
output-relative error rather than first materializing an equivalent dense
matrix.

\subsection{Workload Case-study Evaluation}

The workload case-study artifact is regenerated with \texttt{mise exec -- uv
run python examples/case\_study\_artifacts.py --json-output
docs/results/workload\_case\_studies.json}. It is evidence for provenance
preservation and planner choice, not timing performance. The artifact records a
merged LoRA-style adapter, multi-channel \texttt{nn.Conv1d}, functional
fixed-bias \texttt{conv1d}, grouped \texttt{nn.Conv1d}, depthwise functional
\texttt{conv1d}, a fixed causal-band mask represented as a sparse linear map,
and a fixed per-tensor affine quantized linear module. For each row, the public
fields include the captured operator kind, shape, lowerings, provenance notes,
selected lowering, output-relative error against the reference workload, dense
fallback lowering, planner cost proxy, memory proxy, and an explicit
\texttt{measured\_timing: false} boundary.

The adapter case shows a provenance-only planner opportunity. The Torch module
executes with a merged dense adapter weight, but the capture pass can associate
the nearby \texttt{lora\_A}/\texttt{lora\_B} factors with that merged weight.
With that provenance preserved, the planner selects
\texttt{low\_rank\_product\_bias}; after dense materialization, the valid
fallback is \texttt{dense\_gemm\_bias}. The case-study row checks the selected
operator against the Torch module with near-zero output-relative error and
records the same planner proxy fields for the dense fallback. This supports the
claim that adapter provenance changes the available lowering choices even when
the numerical dense matrix is still a valid fallback. It does not prove a
hardware speedup or broad adapter coverage beyond the named-factor pattern
captured by the frontend.

The Conv1d rows exercise a different form of lost structure. Module,
functional, grouped, and depthwise cases preserve channel, group, stride,
padding, dilation, bias, and output-shape provenance before the operation can
be flattened into an equivalent dense matrix. The planner therefore selects
\texttt{conv1d\_channel\_direct},
\texttt{conv1d\_channel\_direct\_bias},
\texttt{conv1d\_grouped\_direct\_bias}, or
\texttt{conv1d\_depthwise\_direct} as appropriate, while dense materialization
keeps \texttt{dense\_gemm} or \texttt{dense\_gemm\_bias} available. These rows
support the narrower claim that preserved convolution provenance exposes
direct channel-aware lowerings and exact dense fallbacks for the implemented
1D cases. They do not establish Conv2d support, non-scalar Conv1d parameter
coverage, quantized convolution contracts, or measured runtime behavior.

\subsection{PEFT Capstone Evaluation}

The external PEFT plus Transformers artifact is regenerated with \texttt{mise
exec -- uv run --with transformers --with accelerate --with safetensors --with
huggingface\_hub python benchmarks/peft\_transformers\_lora\_inference.py
--json-output docs/results/peft\_transformers\_lora\_inference.json}. The
committed artifact from issue \#80 and PR \#90 records mode
\texttt{real}, the contract-defined
\texttt{hf-internal-testing/tiny-random-OPTForCausalLM} base model,
\texttt{peft-internal-testing/tiny-OPTForCausalLM-lora} adapter, CPU fp32
prefill workload, upstream PEFT revision
\texttt{8d5f7842d9178e51292c2be551e1147ad6f0b5f8}, PEFT fork revision
\texttt{7ac8d57b100846837c5a3b76c65e1e1954ccc3c8}, and Transformers
\texttt{5.13.0}.

The result is bounded. The artifact includes all required rows and correctness
passes for seq16 and seq64 across upstream unmerged PEFT, upstream merged
dense, and the Beyond Matmul PEFT fork. On successful fork rows, the fork emits
\texttt{beyond\_matmul\_lora\_provenance} events with rank, factor shapes,
adapter identity, and explicit dense-fallback availability; this supports the
claim that the external stack can expose fixed LoRA provenance without removing
dense fallback. Both seq128 shapes fail across all baselines with
\texttt{index out of range in self}, so the artifact records
\texttt{summary.benchmark\_ready: false},
\texttt{readiness\_blockers: ["correctness\_checks\_failed"]}, and
\texttt{summary.performance\_claim: "none"}. The fork is not claimed to be
faster than upstream PEFT, peak memory is not measurable on the CPU run, and
adapter switching is \texttt{not\_measured\_single\_adapter}.

The fixed-mask row is intentionally narrower than full masked attention. It
applies a fixed causal-band binary mask as a sparse linear map and selects
\texttt{fixed\_mask\_sparse}, while dense materialization remains the fallback.
This row supports fixed-mask provenance and fallback comparisons for a
deterministic linear map. It does not claim score-dependent attention masking,
softmax behavior, KV-cache layout support, or general mask-aware attention
coverage.

The quantized-linear row is similarly bounded. A deterministic FX-style capture
preserves the packed integer payload, bit width, integer range, scale, zero
point, and fixed bias for the tested per-tensor affine module pattern. The
planner selects \texttt{packed\_affine\_kernel\_bias}, while dense dequantized
materialization remains available through \texttt{dense\_gemm\_bias}. This row
supports the provenance and fallback claim for the fixed quantized contract; it
does not measure production integer kernels, hardware-specific packing, or
runtime speedup.

The case studies and controlled benchmarks should therefore be read together.
The workload artifact demonstrates where provenance changes the planner's
choice set on small Torch examples. The fixed-weight benchmark artifact is the
source for pure-Python timing proxies. The planner-contract and approximation
artifacts bound the planner and error-contract claims. Future workload
families, including quantized convolutional modules, full attention or
score-dependent mask-aware cases, and broader CNN-style blocks, remain
limitations until their frontend, IR, tests, and artifacts land.

The approximation and planner-contract artifacts bound the negative claims. In
the approximation ablation, low-rank and sparse top-k candidates pass the
matrix-relative threshold but fail the output-relative threshold on the sampled
inputs, so the output-aware planner keeps dense GEMM for that case. In the
planner-contract ablation, exact-only planning keeps dense GEMM, bounded-error
planning can choose low-rank under the stated error contract, reuse sensitivity
changes whether preprocessing-heavy options are valid, and unsupported GPU
specialization falls back to dense GEMM. These are deterministic Python
evidence artifacts. They support claims about planner behavior, proxies, and
case coverage; they do not establish hardware-calibrated speedups or production
backend performance.

\subsection{Recovery and Approximation Evaluation}

The recovery evidence answers a narrower question than frontend preservation:
when only a dense matrix remains, can the system expose plausible structure
without pretending that the original computation is known? The current answer
is yes for selected probes, with explicit uncertainty. Unit tests cover
ambiguous dense matrices whose diagonal, sparse, codebook, and low-rank
candidate confidences remain bounded rather than high-confidence by default.
The fixed-weight demo then exercises the positive path on a materialized
rank-four adapter matrix. Recovery emits low-rank candidates with confidence
$1.0$ and output-relative validation errors at numerical-noise scale on the
provided samples, while a diagonal candidate is rejected by validation with
\texttt{output\_relative\_l2} near $0.996$ and a validation confidence cap of
$0$. These numbers are evidence that confidence is now tied to probe details
and sample validation; they are not calibration data. A different sample set or
unseen deployment distribution can still invalidate a recovered approximation,
so dense GEMM remains the exact fallback.

The approximation ablation isolates why matrix reconstruction error is an
insufficient acceptance rule. On the deterministic
\texttt{dominant\_unused\_feature} case, the request allows approximate
lowerings up to relative error $0.25$. A rank-one candidate has matrix-relative
error $0.0685$ and a sparse top-k candidate has matrix-relative error $0.0688$,
so a reconstruction-only rule would accept both. On the representative inputs,
their output-relative errors are about $0.996$ and $1.0$, respectively, so the
output-aware contract rejects both and selects \texttt{dense\_gemm}. The
codebook and bitpacked candidates fail both matrix and output thresholds. This
artifact supports the planner contract choice to score approximate lowerings
against product or output behavior when samples exist. It does not claim that
low-rank or sparse approximations are generally poor; it shows a small case
where the two error notions lead to different qualitative decisions.

The planner-contract ablation connects those error contracts to lowering
selection. On the same rank-one-plus-noise fixed-weight case, an exact-only
request selects \texttt{dense\_gemm} with zero relative error and an estimated
cost proxy of $208$. A bounded-error request with $\epsilon=0.1$ selects
\texttt{low\_rank\_product} with observed relative error about $0.00235$ and an
estimated cost proxy of $157$, while keeping dense GEMM valid. Reuse contracts
change validity before they change semantics: at seven requested calls the
same preprocessing-heavy low-rank option is rejected because preprocessing does
not amortize, and at eight calls it becomes valid and selected. Backend
contracts are similarly explicit: a GPU request rejects an otherwise exact
\texttt{codebook\_kernel} because the requested backend does not support that
lowering, then selects the dense fallback. These costs are planner estimates in
a deterministic artifact, not measured hardware timings.

\section{Related Work}

Beyond Matmul is closest to work that keeps linear structure visible long
enough to choose a better representation, but it is not claiming that
structured operators, compression, quantization, or compiler metadata are new.
The narrower claim is that a fixed-weight inference system can benefit from
making the reason for a matrix multiplication explicit until planning time.
This first public artifact deliberately treats related work as a scoped map
rather than a complete bibliography: it names the research areas the argument
depends on, records how this artifact differs from them, and leaves canonical
citation curation to a later paper-polish pass that does not change the
executable claims.

\subsection{Structured Linear Operators}

Structured linear operators are a long-running alternative to treating every
linear map as an arbitrary dense matrix. Convolutions, Toeplitz and circulant
maps, diagonal scalings, Kronecker-like factorizations, fast transforms, and
graph or stencil operators all expose algebraic constraints that can reduce
storage or work when the backend supports the corresponding kernel.
This project uses that lineage as motivation, but its current evidence is
smaller: executable dense
equivalence and planner behavior for the operator families listed in the IR
section. It does not claim that every structured family is supported or that a
specialized kernel is always faster.

\subsection{Sparse and Low-Rank Inference}

Sparse inference systems and low-rank adaptation methods show that many
inference workloads can use fewer degrees of freedom than a dense matrix
exposes. Sparse kernels exploit zeros or block patterns, while low-rank
methods preserve a factorized product instead of materializing the merged
dense weight. Beyond Matmul borrows these representations as planner
candidates. The distinction is the contract around provenance and fallback: a
low-rank adapter captured before
densification carries different evidence than a low-rank approximation
recovered from a dense matrix, and dense GEMM remains valid when sparsity or
rank evidence is weak.

\subsection{Quantized and Packed Weights}

Quantization, codebook methods, and packed binary or low-bit weights already
separate storage format from mathematical dense multiplication. Their practical
benefit depends on scale and zero-point semantics, packing layout, calibration
or error bounds, and backend support. The current artifact only models a small
part of that space through codebook, bitpacked, and per-tensor affine integer
operators. The Torch frontend only claims the tested fixed quantized
\texttt{nn.Linear} case; broader quantized Torch modules remain future work
until the IR, frontend, planner, and evidence artifacts can represent their
contracts without silently dequantizing away the relevant structure.

\subsection{Compiler IR and Lowering}

ML compilers routinely preserve information across lowering levels through
operator dialects, layout attributes, shape constraints, buffer metadata, and
target-specific scheduling decisions. Beyond Matmul is compatible with that
tradition rather than a replacement for it. Its prototype IR is intentionally
small and research-facing: it records enough
source, structure, exactness, approximation, reuse, layout, and backend
metadata to test whether fixed-weight planner decisions improve over immediate
dense lowering. Production compiler concerns such as full graph optimization,
autograd, memory planning, and GPU code generation remain outside the current
scope.

\subsection{Provenance and Metadata}

Database systems, compilers, and ML tooling all use provenance or metadata to
explain where values came from, whether transformations are legal, and how
artifacts should be interpreted. This paper uses a narrower phrase,
computation reason provenance, for why a fixed linear operator exists and what
lowerings remain valid. That choice connects related work back to the
project's main limitation: provenance that is preserved by a frontend is
stronger than structure guessed from a dense matrix, and recovered candidates
need explicit uncertainty and
output-level validation before a planner should prefer them over dense GEMM.

\section{Development Methodology}

All code work should be test-driven and issue-driven. Each issue should be
small enough for one PR, labeled for priority, risk, area, kind, and status, and
linked to blockers when necessary. Worker agents implement from fresh worktrees
based on latest main. Reviewer agents independently check correctness, local CI
parity, minimality, and north-star alignment before merging. Retrospective
agents periodically trim duplication, detect stuck PRs, and create corrective
issues.

\section{Limitations and Threats to Validity}

The current evidence is intentionally bounded. Benchmark and workload artifacts
exercise small fixed-weight cases, record planner cost and memory proxies, and
include pure-Python timing fields where the benchmark script measures them.
These artifacts are useful for comparing planner choices and generating paper
figures, but they are not hardware-calibrated speedup evidence and do not
stand in for production kernels. Claims about GPU kernels, production runtime
speedups, or backend-specific scheduling are deliberately out of scope until a
future backend study adds measured runtime evidence.

Recovery after provenance is lost is weaker than preserving provenance before
dense lowering. The analyzer confidence values are probe-specific heuristic
scores, not statistically calibrated probabilities, and output-relative
validation only covers the representative samples supplied to the analyzer. A
recovered low-rank, sparse, diagonal, or codebook candidate can be useful
planner evidence, but it does not prove the original computation reason and it
does not replace the dense fallback. Broader recovery-quality benchmarks and
calibrated confidence models remain future work.

Approximation and planner results should be read as contract evidence rather
than broad model-quality or performance evidence. The approximation ablation is
one deterministic synthetic case showing that matrix reconstruction error and
output-relative error can lead to different accept/reject decisions. The
planner-contract ablation checks exactness, bounded error, reuse
amortization, backend support rejection, and dense fallback preservation on
small deterministic scenarios. These results support the need for explicit
error and planning contracts; they do not show that any approximation family is
generally good or bad, nor that the current cost model is globally optimal.

Quantized and mask-aware claims are also narrow by design. The IR can express
exact codebook payloads, tensor-wide scaled binary payloads, and per-tensor
affine integer payloads, and dense dequantization remains a semantic fallback.
Torch frontend capture is limited to the tested fixed per-tensor affine
quantized \texttt{nn.Linear} case. Quantized convolution, per-axis,
per-channel, activation-dependent, mixed-precision, and blockwise quantization
remain outside the current contract. The quantized workload row records
preserved packed-affine provenance, planner proxies, exact output comparison,
and dense dequantized fallback; it is not evidence for production integer
kernels or backend-specific speedup. Similarly, fixed-mask support covers explicit
binary masks applied as sparse linear maps over values or features; it does
not claim score-dependent masked attention, softmax semantics, KV-cache
layout, broadcast mask contracts, or dynamic sequence behavior.

A final threat is claim drift across the paper, docs, wiki, and executable
artifacts. The project mitigates this by treating
\texttt{docs/evidence\_matrix.md} as the claim-to-evidence map and by keeping
unsupported claims labeled as future work or out of scope. During this pass,
the unsupported high-priority claims found in the matrix were already bounded
by existing prose or tracked follow-up work, so this section does not create a
new follow-up issue by itself.

\section{Completion Criteria}

The project reaches its north star when:

\begin{enumerate}
  \item the whitepaper is internally consistent and all major claims map to
  tests, demos, benchmark artifacts, or clearly labeled future work, with the
  current mapping recorded in \texttt{docs/evidence\_matrix.md};
  \item the codebase covers the selected high-leverage fixed-weight operator
  families with exact equivalence and negative tests;
  \item benchmark outputs are generated as machine-readable artifacts suitable
  for paper figures;
  \item the wiki concisely defines the project, north star, coverage boundary,
  issue loop, and artifact map;
  \item GitHub issues show no unresolved priority-zero or priority-one blockers
  against the paper thesis;
  \item retrospective review finds no major duplication, stale claims, or stuck
  PRs that would change the conclusions.
\end{enumerate}

\section{Completion Audit}

The July 2026 draft now contains the final first-artifact argument: motivation,
scope, provenance-aware IR and contracts, frontend capture, recovery,
approximation builders, planner selection, controlled benchmark evidence,
workload case-study evidence, recovery and approximation evaluation, related
work boundaries, limitations, and conclusion. The claim-to-evidence mapping is
centralized in \texttt{docs/evidence\_matrix.md}; the separate completion note
in \texttt{docs/completion\_audit.md} records the final audit of claims,
evidence, limitations, local validation commands, and remaining optional
follow-up issues.

The live issue audit found no unresolved priority-zero or priority-one blocker
that changes the paper thesis. The first-artifact roadmap tracker is historical
coordination rather than the current roadmap. Quantized workload evidence now
covers the tested fixed per-tensor affine linear path, while quantized
convolution capture remains optional follow-up work. These boundaries are
already reflected in the scope, frontend, IR, evidence, and limitations
sections above.

\section{Capstone Boundaries}

The PEFT plus Transformers capstone proves less than the initial performance
hypothesis, but it usefully narrows the research claim. The supported claim is
that, for the measured fixed-LoRA inference workload, a PEFT fork can preserve
structured low-rank adapter provenance and report dense fallback availability
inside a real Transformers stack while matching upstream outputs on seq16 and
seq64 rows. The unsupported claim is that this is benchmark-ready performance
evidence: seq128 failures keep the artifact below the contract's readiness
bar, and the artifact intentionally records no latency, memory, adapter
switching, training, or universal Transformer speedup conclusion.

\section{Conclusion}

Beyond Matmul treats dense multiplication as a powerful lowering, not the only
truth a compiler or runtime should remember. By preserving computation reason
provenance, the project aims to make structured inference paths measurable,
testable, and available to a planner. The final success condition is not a large
surface area of code; it is a clear research argument with enough executable
evidence to make the argument hard to dismiss.

\end{document}
